---
title: "An Overview of Country Income Classification with Respect to GDP"
author: "Chris Boomhower"
date: "June 18, 2016"
output:
  html_document:
    keep_md: true
---
##Introduction


```{r setup, message=FALSE}
knitr::opts_chunk$set(echo=TRUE)
require(downloader)
require(dplyr)
require(ggplot2)
require(knitr)
```


### GDP Data: Download and Importation

The Gross Domestic Product data set is comprised of 2012 GDP values for various countries throughout the world. More recent data is hosted on Worldbank.org's website at http://data.worldbank.org/data-catalog/GDP-ranking-table.

For this analysis, however, the 2012 data hosted at cloudfront.net is of particular interest. As such, the 2012 data at https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv is downloaded.

```{r}
## Download data
prodURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"

download(prodURL, destfile = "GrossDomesticProduct.csv")

list.files() # Confirm download to working directory
```

<br>
Once the data is dowloaded to R's working directory, it is ready to be imported into the R environment as a data frame. After importation, its internal structure details and beginning and ending rows are observed to determine what actions to take when tidying the data. The last 100 rows from the data are observed since the raw data set contains as many as 95 irrelevant rows in its tail end. These last 100 rows are displayed in table form to shorten the output length.

```{r}
## Import Gross Domestic Product data and review raw
productRaw <- read.csv("GrossDomesticProduct.csv", stringsAsFactors = FALSE, header = FALSE) # Try reading characters in as strings instead of factors for easier manipulation
str(productRaw) # Review raw data internal structure details
head(productRaw) # Review beginning rows to look for irrelevant entries
kable(tail(productRaw, 100), format = "html", caption = "Last 100 Rows of Raw GDP data", align = 'l', row.names = TRUE) # Review ending rows to look for irrelevant entries
```

<br>

### GDP Data: Tidying

Empty rows at the beginning and end of the data are first removed during the tidying process.

```{r}
product <- productRaw[6:236,] # Remove empty rows at beginning and end of productRaw data.frame
str(product) # Review raw data internal structure details once more
head(product) # Review beginning rows once more
tail(product) # Review ending rows once more
```

<br>
Column *V3* and columns *V7* through *V10* are suspected of being empty. The next step is to ensure this is the case by counting the total number of values that are not "NA". If the sum of valid entries between all 5 columns is 0, then the columns may be safely removed from the data.

```{r}
## Check NAs in imported columns to identify empty columns
sum(!is.na(product[,c(3,7:10)])) # Detect total number of valid entries
```

<br>
Since column *V6* imported as a character class, it is suspected to contain potentially valid entries. To confirm this, the sum of rows in column *V6* containing values is compared to the sum of rows in *V6* missing values.

```{r}
## Check character column suspected of having empty character entries
sum(product$V6 != "") # Check character type column for valid entries
sum(product$V6 == "") # Output empty entry counts
```

<br>
Only valid columns are extracted from the imported GDP data based on the NA results discovered previously. After extraction, the columns names are updated to better reflect each variable and its contents.

```{r}
## Extract only valid columns from 'product'
product <- product[,c(1,2,4:6)]

## Provide names for each column
names(product) <- c("Country.Code","Country.Rank", "Economy", "GDP.Millions.of.US.Dollars", "Comments")
names(product) # Ensure names added correctly
```

<br>
The original *Comments* column consists of letter designators which point to a legend (found at the bottom of the raw data set) that contains comments for some observations. In order to make these details accessible for future use after tidying the data, the *Comments* column letter designators are replaced with the original data set's legend contents.

```{r}
## Replace comment reference with comment from original data's legend
product[product$Comments != "",] # View valid comment column entries before edits
product$Comments[product$Comments == 'a'] <- "Includes Former Spanish Sahara"
product$Comments[product$Comments == 'b'] <- "Excludes South Sudan"
product$Comments[product$Comments == 'c'] <- "Covers mainland Tanzania only"
product$Comments[product$Comments == 'd'] <- "Data are for the area controlled by the government of the Republic of Cyprus"
product$Comments[product$Comments == 'e'] <- "Excludes Abkhazia and South Ossetia"
product$Comments[product$Comments == 'f'] <- "Excludes Transnistria"
product[product$Comments != "",] # View valid comment column entries after edits
```

<br>
*GDP.Millions.of.Us.Dollars* was imported into the *product* data frame as a character class variable since it contains commas, which are used as thousands seperators, and periods, which are used to represent unavailable or uncollected data. In order to convert *GDP.Millions.of.Us.Dollars* to a numeric class for further processing, these commas and periods are removed.

```{r}
## Prepare GDP column for type conversion
product$GDP.Millions.of.US.Dollars <- gsub(",","", product$GDP.Millions.of.US.Dollars, fixed = TRUE) # Remove commas (fixed = TRUE to remove "," as is)
product$GDP.Millions.of.US.Dollars <- gsub(".","", product$GDP.Millions.of.US.Dollars, fixed = TRUE) # Remove periods (fixed = TRUE to remove "." as is)
```

<br>
Now that commas and periods are removed, the *GDP.Millions.of.Us.Dollars* column class is converted to numeric. Since no non-numerical characters are present within the *Country.Rank* column, this column class is ready for conversion from character to integer at this time as well.

```{r}
## Convert column classes as necessary
product$GDP.Millions.of.US.Dollars <- as.numeric(product$GDP.Millions.of.US.Dollars) # Convert GDP type to numeric
product$Country.Rank <- as.integer(product$Country.Rank) # Convert Country.Rank type to integer
```

<br>
In order to gain a good perspective as to how many rows are still missing data before merging is performed, all rows still missing data are observed within R.

```{r}
product[!complete.cases(product),] # View all rows for which not all variable data is available
```

<br>
As data will be merged by *Country.Code*, it is important to remove any rows missing a country code. This is done next.

```{r}
product1 <- subset(product, product$Country.Code != "") # Remove missing Country.Code values from the data frame
nrow(product1[!complete.cases(product1$Country.Code),]) # Confirm there are no missing Country.Code values in the data
```

<br>
Relevant GDP data is ready for final extraction in preparation for merging with the income data. In this case, *Country.Code*, *Country.Rank*, and *GDP.Millions.of.Us.Dollars* columns are extracted. To ensure only these data are present, the internal structure details are reviewed. Finally, the data is ordered by *Country.Code*

```{r}
GDPdata <- product1[,c(1,2,4)] # Extract only Country.Code, Country.Rank and GDP column data to be merged
str(GDPdata) # Review extracted data internal structure details

GDPdata <- GDPdata[order(GDPdata$Country.Code),] # Order the data by Country.Code instead of GDP

## NOTE: TO WRITE THIS CLEANED DATA SET TO CSV, UNCOMMENT THE FOLLOWING LINE OF CODE
# write.csv(GDPdata, "Product_clean.csv", row.names = FALSE)

```

<br>

## Education Data: Download and Importation

The Education data set, which includes the country income data of interest, is comprised of details such as education access, completion, literacy, population, teachers, etc. for various countries. More recent data is hosted on Worldbank.org's website at http://data.worldbank.org/data-catalog/ed-stats.

For this analysis, the most recently updated copy of the data is downloaded from https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv.

```{r}
## Download data
educURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"
download(educURL, destfile = "Education.csv")

list.files() # Confirm download to working directory
```

<br>
Once this second data set is dowloaded to R's working directory, it is ready to be imported into the R environment as a data frame, just as was done for the GDP data. Once again, after importation its internal structure details and beginning and ending rows are observed to determine what actions to take when tidying the data.

```{r}
## Import Educational data and review raw data
EducRaw <- read.csv("Education.csv", stringsAsFactors = FALSE, header = TRUE) # Try reading characters in as strings instead of factors for easier manipulation
str(EducRaw) # Review raw data internal structure details
head(EducRaw) # Review beginning rows to look for irrelevant entries
tail(EducRaw) # Review ending rows to look for irrelevant entries
```

<br>

### Education Data: Tidying

As the country code columns will later be matched between GDP and income data, the name *CountryCode* is updated to *Country.Code* to reflect the same name used in the GDP data. All remaining column names are acceptable.

```{r}
## Rename CountryCode variable to match GDPdata's Country.Code
Education <- rename(EducRaw, Country.Code = CountryCode)
```

<br>
Since only the *Country.Code*, *Income.Group*, and *Short.Name* columns will be used in the final analysis, the number of NA's present in each of these columns is observed before merging these data with the GDP data.

```{r}
## Check NAs or missing values in columns of interest
nrow(Education[Education$Country.Code == "",])
nrow(Education[Education$Income.Group == "",])
nrow(Education[Education$Short.Name == "",])
```

<br>
*Country.Code*, *Income.Group*, and *Short.Name* columns are finally extracted in preparation for merging. One last check is performed to ensure there are no missing entries in the *Country.Code* column.

```{r}
## Extract CountryCode and Income.Group columns
Income <- Education[,c(1,3,31)]
head(Income)

nrow(Income[Income$Country.Code == "",]) # Confirm there are no missing Country.Code values in the data

## NOTE: TO WRITE THIS CLEANED DATA SET TO CSV, UNCOMMENT THE FOLLOWING LINE OF CODE
# write.csv(Income, "Income_clean.csv", row.names = FALSE)

```

<br>

### Merging GDP and Income Data

Now that the original data have been cleaned and only the columns of interest have been set aside for merging, GDP and Income data are ready to be merged together. After merging the two data sets together by *Country.Code*, the merged data frame's internal structure details and beginning and ending rows are observed to make sure the data merged correctly.
```{r}
## Merge Income and GDPdata
MergeData <- merge(Income, GDPdata, by = "Country.Code", all = TRUE)
str(MergeData) # Review raw data internal structure details
head(MergeData) # Review beginning rows to ensure no blank observations
```

<br>

### Question 1: Match the data based on the country shortcode. How many of the IDs match?

```{r}
## Indicate how many of the IDs match
length(intersect(GDPdata$Country.Code, Income$Country.Code))

## Indicate how many of the rows contain NAs
sum(!complete.cases(MergeData))

## Remove rows with missing data
MergeData1 <- MergeData[complete.cases(MergeData),]
nrow(MergeData1) # Provide row count after removing rows with missing data
```

<br>

### Question 2: Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?

```{r}
MergeData1 <- MergeData1[order(MergeData1$GDP.Millions.of.US.Dollars, decreasing = FALSE),] # Sort the data frame by GDP
MergeData1$Short.Name[13] # Display only the 13th country in the data frame
```

<br>

### Question 3: What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?

```{r}
mean(subset(MergeData1, Income.Group == "High income: OECD")$Country.Rank)
mean(subset(MergeData1, Income.Group == "High income: nonOECD")$Country.Rank)
```

<br>

### Question 4: Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income.Group

```{r}
# MergeData1$Country.Code <- as.factor(MergeData1$Country.Code) # Convert Country.Code type to factor in order to reorder ggplot x-axis by Income.Group
# MergeData1$Country.Code <- factor(MergeData1$Country.Code, levels = MergeData1$Country.Code[order(MergeData1$GDP.Millions.of.US.Dollars)]) # Reorder Country.Code by Income.Group

ggplot(data = MergeData1, aes(x=Income.Group, y=GDP.Millions.of.US.Dollars, fill=Income.Group)) +
    geom_boxplot() +                                                                            # Create boxplots
    theme(axis.text.x = element_text(angle = 55, hjust = 1, size = 10)) +                       # Adjust X axis label size and position
    xlab("Income Group") + ylab("GDP (Millions of US Dollars)") + ggtitle("GDP for All Countries by Income Group") # Provide labels

ggplot(data = MergeData1, aes(x=Income.Group, y=log(GDP.Millions.of.US.Dollars), fill=Income.Group)) + # Re-plot log transformed GDP
    geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=3, fill="red") +     
    geom_jitter(shape=16, position=position_jitter(0.3)) +                                      # Add individual data points with jitter
    theme(axis.text.x = element_text(angle = 55, hjust = 1, size = 10)) +                       
    xlab("Income Group") + ylab("Log Transformed GDP (Millions of US Dollars)") + ggtitle("Log Transformed GDP for All Countries by Income Group") # Provide labels
```

<br>

### Question 5: Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are lower middle income but among the 38 nations with the highest GDP?

```{r}
## Create additional data frame and add quantile column
MergeData2 <- MergeData1
MergeData2$GDP.Quantile <- ntile(MergeData2$Country.Rank, 5) # Add 5 quantiles by Country.Rank to new GDP.Quantile column

## Generate table by Income.Group output
table(MergeData2$Income.Group, MergeData2$GDP.Quantile, dnn = c("Income.Group","GDP.Quantile"))

## Confirm number of lower middle income countries as listed in the table above
sum(MergeData2[(nrow(MergeData2)-37):nrow(MergeData2),]$Income.Group == "Lower middle income")

```